---
title: "Personal Activity Monitoring Analysis"
author: "Tynisha Niles"
date: "2026-03-23"
output: 
  html_document:
  keep_md: true 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This project analyzes personal activity data collected at 5-minute intervals over two months (October–November 2012). The dataset contains the number of steps taken per interval, allowing us to explore daily activity patterns, missing data, and differences between weekdays and weekends.

---

## Loading and Preprocessing the Data

```{r echo=TRUE}

# Load required libraries
library(dplyr)
library(ggplot2)

# Load dataset
data <- read.csv("activity.csv")
data$date <- as.Date(data$date)

# View structure
str(data)


# -------------------------------
# Mean Total Steps Per Day
# -------------------------------

steps_per_day <- data %>%
  group_by(date) %>%
  summarise(total_steps = sum(steps, na.rm = TRUE))

# Histogram

hist(steps_per_day$total_steps,
     main="Total Steps per Day",
     xlab="Steps",
     col="lightblue",
     breaks=20)

# Mean and Median

mean_steps <- mean(steps_per_day$total_steps)
median_steps <- median(steps_per_day$total_steps)

mean_steps
median_steps

# -------------------------------
# Average Daily Activity Pattern
# -------------------------------

interval_avg <- data %>%
  group_by(interval) %>%
  summarise(avg_steps = mean(steps, na.rm = TRUE))

# Time series plot

plot(interval_avg$interval, interval_avg$avg_steps,
     type="l",
     xlab="5-minute Interval",
     ylab="Average Steps",
     main="Average Daily Activity Pattern")

# Interval with max steps

max_interval <- interval_avg[which.max(interval_avg$avg_steps),]
max_interval

# -------------------------------
# Missing Values

missing_values <- sum(is.na(data$steps))
missing_values

# -------------------------------
# Imputation Strategy
# -------------------------------
interval_means <- interval_avg

data_imputed <- merge(data, interval_means, by="interval")

data_imputed$steps[is.na(data_imputed$steps)] <- 
  data_imputed$avg_steps[is.na(data_imputed$steps)]

data_imputed <- data_imputed[, c("steps", "date", "interval")]

# -------------------------------
# Histogram After Imputation
# -------------------------------

steps_per_day_imputed <- data_imputed %>%
  group_by(date) %>%
  summarise(total_steps = sum(steps))

hist(steps_per_day_imputed$total_steps,
     main="Total Steps per Day (Imputed)",
     xlab="Steps",
     col="lightgreen",
     breaks=20)

# Mean and Median after imputation

mean_imputed <- mean(steps_per_day_imputed$total_steps)
median_imputed <- median(steps_per_day_imputed$total_steps)

mean_imputed
median_imputed

# -------------------------------
# Weekday vs Weekend Analysis
# -------------------------------

data_imputed$day_type <- ifelse(weekdays(data_imputed$date) %in% 
                                 c("Saturday", "Sunday"),
                               "weekend", "weekday")

data_imputed$day_type <- as.factor(data_imputed$day_type)

interval_daytype <- data_imputed %>%
  group_by(interval, day_type) %>%
  summarise(avg_steps = mean(steps))

ggplot(interval_daytype, aes(x=interval, y=avg_steps)) +
  geom_line() +
  facet_wrap(~day_type, ncol=1) +
  labs(title="Activity Patterns: Weekday vs Weekend",
       x="Interval",
       y="Average Steps")

```

##Conclusion

Daily step counts vary across days.
A specific 5-minute interval shows peak activity.
Missing values affect total step calculations.
Imputation improves estimates.
Weekday and weekend activity patterns differ.